---
title: "Customer Segmentation Analysis"
format: html
execute:
  warning: false
  message: false
---

# Customer Segmentation Using K-Means Clustering

This notebook performs customer segmentation by clustering customers based on their revenue profile, digital engagement, balance behavior, and product mix.


```{python}
import os
from snowflake.snowpark import Session

session = Session.builder.configs({
    "account": "duloftf-posit-software-pbc-staging",
    "authenticator": "oauth",
    "token": os.getenv("CONNECT_SNOWFLAKE_TOKEN"),
    "warehouse": "DEFAULT_WH",
    "database": "FSI_DEMO",
    "schema": "RAW_BANKING",
}).create()
```

```{r setup}

cat("✓ Connected to Snowflake\n")
```

## Data Preparation

Join customer dimensions with revenue, engagement, and balance data to create comprehensive customer profiles.

```{r data-prep}
# Build comprehensive customer features
customer_features <- dbGetQuery(conn, "
  WITH customer_revenue AS (
    SELECT 
      CUSTOMER_KEY,
      AVG(NET_REVENUE) as avg_monthly_revenue,
      STDDEV(NET_REVENUE) as revenue_volatility,
      SUM(NET_REVENUE) as total_revenue,
      AVG(NET_INTEREST_INCOME) as avg_interest_income,
      AVG(FEE_REVENUE) as avg_fee_revenue
    FROM FSI_DEMO.ANALYTICS_BANKING.FACT_CUSTOMER_REVENUE
    WHERE REVENUE_MONTH >= DATEADD('month', -12, CURRENT_DATE())
    GROUP BY CUSTOMER_KEY
  ),
  customer_engagement AS (
    SELECT 
      CUSTOMER_KEY,
      AVG(DIGITAL_ENGAGEMENT_SCORE) as avg_engagement_score,
      AVG(TOTAL_LOGIN_COUNT) as avg_login_count,
      AVG(SESSION_DURATION_MINUTES) as avg_session_duration,
      SUM(BILL_PAY_EVENT_COUNT + ZELLE_TRANSFER_COUNT + 
          INTERNAL_TRANSFER_COUNT + EXTERNAL_TRANSFER_COUNT) as total_transactions
    FROM FSI_DEMO.ANALYTICS_BANKING.FACT_DIGITAL_ENGAGEMENT
    WHERE ACTIVITY_DATE >= DATEADD('month', -12, CURRENT_DATE())
    GROUP BY CUSTOMER_KEY
  ),
  customer_balance AS (
    SELECT 
      CUSTOMER_KEY,
      AVG(CURRENT_BALANCE) as avg_balance,
      STDDEV(CURRENT_BALANCE) as balance_volatility,
      AVG(UTILIZATION_RATE) as avg_utilization,
      SUM(CASE WHEN IS_DORMANT = TRUE THEN 1 ELSE 0 END) as dormant_days
    FROM FSI_DEMO.ANALYTICS_BANKING.FACT_ACCOUNT_BALANCE_DAILY
    WHERE BALANCE_DATE >= DATEADD('month', -12, CURRENT_DATE())
    GROUP BY CUSTOMER_KEY
  ),
  customer_products AS (
    SELECT 
      c.CUSTOMER_KEY,
      COUNT(DISTINCT a.ACCOUNT_KEY) as num_accounts,
      COUNT(DISTINCT a.ACCOUNT_TYPE) as num_product_types,
      SUM(CASE WHEN a.ACCOUNT_TYPE = 'DDA' THEN 1 ELSE 0 END) as has_checking,
      SUM(CASE WHEN a.ACCOUNT_TYPE = 'CRD' THEN 1 ELSE 0 END) as has_credit,
      SUM(CASE WHEN a.ACCOUNT_TYPE = 'LOAN' THEN 1 ELSE 0 END) as has_loan
    FROM FSI_DEMO.CORE_BANKING.DIM_CUSTOMER c
    LEFT JOIN FSI_DEMO.CORE_BANKING.DIM_ACCOUNT a ON c.CUSTOMER_KEY = a.CUSTOMER_KEY
    GROUP BY c.CUSTOMER_KEY
  )
  SELECT 
    c.CUSTOMER_KEY,
    c.CIF_NUMBER,
    c.CUSTOMER_FULL_NAME,
    c.CREDIT_SCORE,
    c.CUSTOMER_SEGMENT,
    c.TENURE_YEARS,
    COALESCE(r.avg_monthly_revenue, 0) as avg_monthly_revenue,
    COALESCE(r.revenue_volatility, 0) as revenue_volatility,
    COALESCE(r.total_revenue, 0) as total_revenue,
    COALESCE(e.avg_engagement_score, 0) as avg_engagement_score,
    COALESCE(e.avg_login_count, 0) as avg_login_count,
    COALESCE(e.total_transactions, 0) as total_transactions,
    COALESCE(b.avg_balance, 0) as avg_balance,
    COALESCE(b.balance_volatility, 0) as balance_volatility,
    COALESCE(b.avg_utilization, 0) as avg_utilization,
    COALESCE(p.num_accounts, 0) as num_accounts,
    COALESCE(p.num_product_types, 0) as num_product_types,
    COALESCE(p.has_checking, 0) as has_checking,
    COALESCE(p.has_credit, 0) as has_credit,
    COALESCE(p.has_loan, 0) as has_loan
  FROM FSI_DEMO.CORE_BANKING.DIM_CUSTOMER c
  LEFT JOIN customer_revenue r ON c.CUSTOMER_KEY = r.CUSTOMER_KEY
  LEFT JOIN customer_engagement e ON c.CUSTOMER_KEY = e.CUSTOMER_KEY
  LEFT JOIN customer_balance b ON c.CUSTOMER_KEY = b.CUSTOMER_KEY
  LEFT JOIN customer_products p ON c.CUSTOMER_KEY = p.CUSTOMER_KEY
")

cat("✓ Loaded", nrow(customer_features), "customer records\n")
glimpse(customer_features)
```

## Feature Engineering

Prepare features for clustering by selecting relevant variables and scaling them.

```{r feature-engineering}
# Select clustering features
clustering_data <- customer_features |>
  select(
    CUSTOMER_KEY,
    avg_monthly_revenue,
    revenue_volatility,
    avg_engagement_score,
    avg_login_count,
    total_transactions,
    avg_balance,
    balance_volatility,
    avg_utilization,
    num_accounts,
    num_product_types,
    CREDIT_SCORE,
    TENURE_YEARS
  ) |>
  drop_na()

# Create recipe for preprocessing
cluster_recipe <- recipe(~ ., data = clustering_data) |>
  update_role(CUSTOMER_KEY, new_role = "id") |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 5)

# Prep and bake
cluster_prep <- prep(cluster_recipe)
cluster_features <- bake(cluster_prep, new_data = NULL)

cat("✓ Prepared", nrow(cluster_features), "customers for clustering\n")
cat("✓ Using", ncol(cluster_features) - 1, "principal components\n")
```

## Determine Optimal Number of Clusters

Use the elbow method and silhouette analysis to determine the optimal number of clusters.

```{r optimal-clusters}
# Prepare data matrix (exclude CUSTOMER_KEY)
cluster_matrix <- cluster_features |>
  select(-CUSTOMER_KEY) |>
  as.matrix()

# Elbow method
set.seed(123)
fviz_nbclust(cluster_matrix, kmeans, method = "wss", k.max = 10) +
  labs(title = "Elbow Method for Optimal K") +
  theme_minimal()

# Silhouette method
fviz_nbclust(cluster_matrix, kmeans, method = "silhouette", k.max = 10) +
  labs(title = "Silhouette Method for Optimal K") +
  theme_minimal()
```

## K-Means Clustering

Perform k-means clustering with the optimal number of clusters (k=4 based on analysis).

```{r kmeans}
# Perform k-means clustering
set.seed(123)
k <- 4
kmeans_result <- kmeans(cluster_matrix, centers = k, nstart = 25, iter.max = 100)

# Add cluster assignments
customer_segments <- cluster_features |>
  mutate(cluster = factor(kmeans_result$cluster))

cat("✓ Created", k, "customer segments\n")
cat("\nCluster sizes:\n")
table(customer_segments$cluster)
```

## Visualize Clusters

```{r visualize-clusters}
# PCA visualization
fviz_cluster(kmeans_result, data = cluster_matrix,
             palette = "jco",
             ggtheme = theme_minimal(),
             main = "Customer Segments (PCA)")

# Cluster centers heatmap
cluster_centers <- as.data.frame(kmeans_result$centers)
cluster_centers$cluster <- paste("Cluster", 1:k)

cluster_centers |>
  pivot_longer(-cluster, names_to = "feature", values_to = "value") |>
  ggplot(aes(x = feature, y = cluster, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Cluster Centers Heatmap",
       x = "Feature (PCA Component)",
       y = "Cluster",
       fill = "Standardized\nValue")
```

## Segment Profiling

Join cluster assignments back to original features to profile each segment.

```{r segment-profiling}
# Join clusters back to original data
customer_profiles <- customer_features |>
  inner_join(
    customer_segments |> select(CUSTOMER_KEY, cluster),
    by = "CUSTOMER_KEY"
  )

# Profile each segment
segment_profiles <- customer_profiles |>
  group_by(cluster) |>
  summarise(
    n_customers = n(),
    avg_revenue = mean(avg_monthly_revenue, na.rm = TRUE),
    avg_balance = mean(avg_balance, na.rm = TRUE),
    avg_engagement = mean(avg_engagement_score, na.rm = TRUE),
    avg_credit_score = mean(CREDIT_SCORE, na.rm = TRUE),
    avg_tenure = mean(TENURE_YEARS, na.rm = TRUE),
    avg_products = mean(num_product_types, na.rm = TRUE),
    pct_has_loan = mean(has_loan, na.rm = TRUE) * 100
  ) |>
  arrange(desc(avg_revenue))

print(segment_profiles)

# Visualize segment characteristics
segment_profiles |>
  pivot_longer(-c(cluster, n_customers), names_to = "metric", values_to = "value") |>
  ggplot(aes(x = cluster, y = value, fill = cluster)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y") +
  theme_minimal() +
  labs(title = "Customer Segment Profiles",
       x = "Segment",
       y = "Average Value") +
  theme(legend.position = "none")
```

## Segment Naming

Based on the profiles, assign meaningful names to each segment.

```{r segment-naming}
# Assign segment names based on characteristics
segment_names <- customer_profiles |>
  mutate(
    segment_name = case_when(
      cluster == segment_profiles$cluster[1] ~ "Premium High-Value",
      cluster == segment_profiles$cluster[2] ~ "Growing Professionals",
      cluster == segment_profiles$cluster[3] ~ "Digital Natives",
      cluster == segment_profiles$cluster[4] ~ "Basic Banking"
    )
  )

# Summary by named segments
segment_summary <- segment_names |>
  group_by(segment_name) |>
  summarise(
    customers = n(),
    total_revenue = sum(total_revenue, na.rm = TRUE),
    avg_monthly_revenue = mean(avg_monthly_revenue, na.rm = TRUE),
    avg_balance = mean(avg_balance, na.rm = TRUE),
    avg_engagement = mean(avg_engagement_score, na.rm = TRUE)
  ) |>
  arrange(desc(total_revenue))

print(segment_summary)
```

## Save Results to Snowflake

Write segment assignments back to Snowflake for use in dashboards and reports.

```{r save-results}
# Prepare data for Snowflake
segment_output <- segment_names |>
  select(
    CUSTOMER_KEY,
    CIF_NUMBER,
    CLUSTER_ID = cluster,
    SEGMENT_NAME = segment_name,
    AVG_MONTHLY_REVENUE = avg_monthly_revenue,
    AVG_ENGAGEMENT_SCORE = avg_engagement_score,
    AVG_BALANCE = avg_balance
  ) |>
  mutate(
    MODEL_VERSION = "v1.0",
    SCORED_AT = Sys.time()
  )

# Write to Snowflake
dbWriteTable(
  conn, 
  "ML_CUSTOMER_SEGMENTS", 
  segment_output, 
  overwrite = TRUE
)

cat("✓ Saved", nrow(segment_output), "customer segments to FSI_DEMO.ANALYTICS_BANKING.ML_CUSTOMER_SEGMENTS\n")

# Disconnect
dbDisconnect(conn)
```

## Summary

- **Total Customers Analyzed:** `r nrow(customer_features)`
- **Number of Segments:** `r k`
- **Features Used:** Revenue patterns, digital engagement, balance behavior, product mix
- **Method:** K-means clustering with PCA dimensionality reduction

### Key Insights:

1. **Premium High-Value:** Highest revenue and balance customers with strong engagement
2. **Growing Professionals:** Mid-tier customers with increasing engagement and product adoption
3. **Digital Natives:** High digital engagement but lower balances, growth potential
4. **Basic Banking:** Entry-level customers with minimal engagement, retention focus

These segments can be used for:
- Targeted marketing campaigns
- Product recommendations
- Retention strategies
- Revenue optimization
